---
title: "NFL Data Exploration"
date: "2026-02-01"
format: 
  pdf
html-math-method: mathjax
execute:
  eval: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(here)
library(dplyr)

```

## Background of NFL Data

-   ***Prediction of Retirement for a Quarterback***

    -   **Event:** Retirement (Last game/season in NFL)

    -   **Time Scale:** Number of Seasons/Games since debut

    -   **Censoring:** QB still active at end of observation period

-   **Source:** [NFL Statistics](https://www.kaggle.com/datasets/kendallgillies/nflstatistics/data) (scraped from official NFL website)

-   This data stops at 2016!

-   Relevant Files Variables:

    1.  Basic_Stats

        1.   Name/Player Id (not equal amount)
        2.  Age/Birthday (use the one with less NA?), Height, Weight
        3.  Current Status, Current Team (alot NA), Position (82% NA)
        4.  Experience calculated from Years Played, Years Played (18% null)

    2.  Game_Logs_Quarterback

        Since so many positions are NA, just use Name/Player Id and this file to identify the QBs.

        Filter using Games Played

        1.  restrict to just regular season (82%) ???

        2.  Year, Season, Week, Game Date

        3.  home/away (50/50 split), Outcome

        4.  Games Played 0/1, Passed Completed, Passed Attempted, Completion Percentage, Passing Yards

        5.  Yards Per Carry (Rushing Yards over Attempts, 62% NA)

        6.  Sacks, Sacked Yards Lost, Interceptions, TD Passes (50% NA)

        7.  Fumbles (88% NA)

    3.  Career_Stats_Passing (overlap with variables above) - but this is CAREER STATS for each year - already accumulated compared to game logs

        1.  Games Played (Filter if not enough games played that season/year)

        2.  Interception Rate

    4.  Career_Stats_Rushing

        1.  Rushing Yards Per Game

    5.  Career_Stats_Fumbles

        1.  Fumbles Lost

## Import Data

```{r}
raw_player <- read_csv(here("nfl", "data", "Basic_Stats.csv"))
raw_qb_logs <- read_csv(here("nfl", "data", "Game_Logs_Quarterback.csv"))
# very similar data as qb logs, but qb has a bit more 
raw_passes <- read_csv(here("nfl", "data", "Career_Stats_Passing.csv"))
#raw_rushing <- read_csv(here("nfl", "data", "Career_Stats_Rushing.csv"))
#raw_fumbles <- read_csv(here("nfl", "data", "Career_Stats_Fumbles.csv"))
```

## Read Data

```{r}
# Observe the data 
head(raw_player)
head(raw_qb_logs)
head(raw_passes)

```

Key Limitations:

-   Retirement is inferred, not observed

    -   No games logged after season t

    -   No reappearance in future seasons

-   Watch for potential misclassification ( like temporary exits for injury, etc)

-   Fix using censoring –\> censor recent seasons (this data goes up to 2016 so maybe we can disregard 2015 up)

## Cleaning the Data 

### Filter Files for Statistics only on Quarterbacks 

#### *Basic_Stats File*

```{r}
# get distinct QB player IDs
qb_id <- raw_qb_logs |>
  distinct(`Player Id`)

# Clean player stats - ONLY QUARTERBACKS
player_clean <- raw_player |>
  semi_join(qb_id, by = "Player Id") |> 
  select(`Player Id`, Name, Age,`Height (inches)`, `Weight (lbs)`, 
         Experience, `Years Played`) |>
  arrange(Name)
player_clean

# we can possibly calculate debut year using birth year - parse the birthday 
# (if the football era is impactful (competition)

```

#### *Game_Logs_Quarterback File*

```{r}
# Game logs - regular season only
# if any empty slots, replace with 0 for easy calculation of totals 
qb_regular <- raw_qb_logs |> 
  filter(Season == "Regular Season") |>
  group_by(`Player Id`) |>
  mutate(across(everything(), ~str_replace_all(., "--", "0"))) |>
  mutate(Year = as.numeric(Year)) |>
  ungroup()

#desired variables 
# TD-Int Ratio -> Efficiency 
# Sacks -> injury risk -> early retirement 
# run-pass-ratio (rushing yards/ passing yards) testing if mobile QBs retire earlier
# run_pass_ratio = rushing_yards / passing_yards
variables <- c( "Passes Completed", "Passes Attempted", "Completion Percentage",
                "Passing Yards", "Sacks", "Ints", "TD Passes", "Rushing Yards")

qb_career_summary <- qb_regular |>
  select(-c(Week, `Passer Rating`)) |>
  mutate(across(all_of(variables), as.numeric)) |>
  group_by(`Player Id`, Name) |>
  
  summarise(
    # Calculate career timeline for each QB
    First_Year = min(Year, na.rm = TRUE),
    Last_Year  = max(Year, na.rm = TRUE),
    Total_Seasons = length(unique(Year)),
    Total_Games = sum(`Games Played` == 1, na.rm = TRUE),

    # Calculate Totals for Career 
    across(all_of(variables[variables != "Completion Percentage"]),
           sum, na.rm = TRUE),
    .groups = "drop") |>
  
  mutate(
    Time  = Last_Year - First_Year + 1, # survival time in seasons 
    
    # introduce censoring?? the last year is 2016 
    # so maybe censor players active post-2015 are censored
    # Adjust this threshold based on your data's latest year
    # if retired or not (0/1)
    Event = if_else(Last_Year >= 2015, 0, 1),

    # efficiency
    TD_INT = `TD Passes` / pmax(Ints, 1), # row wise math
    
    RUN_PASS = `Rushing Yards` / pmax(`Passing Yards`, 1),

    # categorize career length
    Career_Length = cut(Total_Seasons,
                        breaks = c(0, 2, 5, 10, Inf),
                        labels = c("1-2 seasons", "3-5 seasons", "6-10 seasons", "10+ seasons")))

qb_career_summary
# Note: 18 week seasons usually 

```

Notes:

-   run-pass-ratio

    -   Higher = more mobility, usually more hits from defenders, higher injury risk potentially

        -   Indication: earlier retirement

-   touchdown-interception-ratio

    -   Higher = more TD, better decision-making, protects the ball

        -   Indication: longer career

#### Join Basic_Stats and QB_Logs File

```{r}
# keep everything in QB Logs
qb_combined1 <- qb_career_summary |>
  left_join(player_clean, by = "Player Id")

# all say same thing: "total_seasons", "time", "Experience", "career_length", "Years Played"

# reorder 
order <- c("Player Id", "Name.x", "Event", "Age", "Height (inches)", "Weight (lbs)",
           "First_Year", "Last_Year", "Time",
           "Total_Games", "Passes Completed", "Passes Attempted", "Passing Yards", 
           "Sacks", "Ints", "TD Passes", "Rushing Yards", "TD_INT", "RUN_PASS")

qb_combined2 <- qb_combined1 |>
  select(all_of(order)) |>
  rename(Name = Name.x,
         Retired = Event, 
         Experience = Time)
         
qb_combined2

```

#### Export Clean Data: 

```{r}
safe_write_csv <- function(data, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(data, path)}

safe_write_csv(qb_combined2, here("nfl", "data","cleaned", "1_cleaned_nfl_data.csv"))
```

#### *Career_Stats_Passing File (might not use, similar to QB logs))* 

```{r, eval= FALSE}
# These are YEARLY career stats

#passes <- raw_passes |>
#  semi_join(qb_id, by = "Player Id") |>
#  arrange(Name)
#passes
```

## Explore Data / Missingness

```{r}

```
