---
title: "Data Documentation"
subtitle: "NFL Quarterback Retirement Research"
date: "2026-02-25"
format: 
  pdf
html-math-method: mathjax
execute:
  eval: TRUE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
```

# Research Overview

**Purpose:** Survival Analysis of Quarterback retirement in National Football League (NFL).

**Built by:** `00_nfl_EDA.Rmd` → **`nfl/data/cleaned/5_ALL_DATA.csv`**.

-   **Event:** Retirement = last season in NFL (no games after season *t*).
-   **Time:** Seasons since debut (`Career_Time`).
-   **Censoring:** Active through end of data → `Retired = 0`. Data **stops at 2016**; \~98 QBs censored (EDA).

**Raw inputs:** Basic_Stats.csv, Game_Logs_Quarterback.csv (regular season only), Career_Stats_Passing.csv **Source:** [Kaggle NFL Statistics](https://www.kaggle.com/datasets/kendallgillies/nflstatistics/data); [article](https://fonseca-carlos.medium.com/predicting-qb-retirement-with-survival-analysis-or-can-data-tell-if-brady-will-play-till-50-5d0c4b8c24e9).

# Dataset:

**File:** `5_ALL_DATA.csv`

**Unit:** One row per QB. **Scope:** QBs with ≥1 row in Game_Logs_Quarterback; no minimum games in final file (0-game QBs included). **Merge:** Career + debut-year + first-three-year summaries; no raw career totals as predictors (avoids leaking survival time).

## Load Data and Dimensions

```{r load data}
nfl_data <- read_csv(
  here("nfl", "data", "cleaned", "5_ALL_DATA.csv"),
  show_col_types = FALSE)
```

```{r dimensions}
dim(nfl_data)
# 35 columns, 466 rows (Quarterbacks)
```

# Variable Reference

### *Identifiers and Outcome*

| Variable | Type | Description |
|----|----|----|
| `Player_Id` | character | Unique Player ID. |
| `Name` | character | Player name. |
| `Retired` | 0/1 | **Outcome.** 1 = last season \< 2016; 0 = censored (active through 2016). |

### *Demographics and Career Frame*

+-------------------+---------+-------------------------------------------------------------------------------+
| Variable          | Type    | Description / formula                                                         |
+===================+=========+===============================================================================+
| `Debut_Age`       | numeric | Age at debut: `Age - (2016 - First_Year)` (Original Data: Age = age in 2016). |
+-------------------+---------+-------------------------------------------------------------------------------+
| `Height (inches)` | numeric | Height of Player in inches.                                                   |
+-------------------+---------+-------------------------------------------------------------------------------+
| `Weight (lbs)`    | numeric | Weight of Player in pounds.                                                   |
+-------------------+---------+-------------------------------------------------------------------------------+
| `First_Year`      | numeric | Debut year (\~1970–2016).                                                     |
+-------------------+---------+-------------------------------------------------------------------------------+
| `Career_Time`     | numeric | **Survival time:** seasons from debut = `Last_Year - First_Year + 1`.         |
|                   |         |                                                                               |
|                   |         | Ex. 11 seasons - 2001 to 2011                                                 |
+-------------------+---------+-------------------------------------------------------------------------------+

### *Game counts (ONLY Regular-Season)*

| Variable           | Description                     |
|--------------------|---------------------------------|
| `Career_Games`     | Total Games Played in Career.   |
| `Debut_Year_Games` | Total Games in debut season.    |
| `Three_Year_Games` | Total Games in first 3 seasons. |

### *Derived ratios (Career\_\*\| Debut\_\* \| Three_Year\_\*)*

-   **TD_INT:** `TD Passes / max(Ints, 1)` - Touchdown to Interception ratio; Infer accuracy/efficiency; higher → longer career (EDA).
-   **RUN_PASS:** `Rushing Yards / max(Passing Yards, 1)` — mobility; can be \<0 (sacks); higher → more hits, earlier retirement (EDA).

### *Completion and Per-Game stats (\*\_PG)*

-   **Completion_Percentage** (career / debut / three-year): Number of Throws Completions / Number of Throws Attempts; Infer accuracy/efficiency
-   **Per-Game Statistics:** `*_PG`, `Debut_*_PG`, `Three_Year_*_PG` for Yards (passing), Sacks, TD (touchdowns), Ints (interceptions), Rush (rushing yards).
    -   **Denominator** = Career_Games, Debut_Year_Games, or Three_Year_Games;
        -   **NA when denominator = 0.**

# Missingness and Ranges

**Missingness:** Debut stats have more NAs (no debut games or 0 attempts). Three-year window has fewer NAs. Career stats NA when Total_Games = 0.

-   Most missingness is 'Debut_Completion_Percentage' with 181 of 466 rows NA.

```{r NAs}
missingness <- colSums(is.na(nfl_data))
missingness[missingness > 0]
max(missingness) 
```

**Ranges (EDA):**

-   First_Year (Debut Year): 1970–2016

-   RUN_PASS can be negative

-   Many zeros in TD_INT / per-game for short careers. Consider filtering e.g. `Big_T > 2` or `Total_Games > 0` for modeling.

# Data Structure and Summary

```{r glimpe at structure}
glimpse(nfl_data)
```

```{r summary}
summary(nfl_data)
```

# Limitations

| Topic | Note |
|----|----|
| **Retirement** | Inferred (no games after season *t*). Misclassification possible (injury, etc.). |
| **Censoring** | Cutoff 2016; active QBs = 0. Consider sensitivity to cutoff. |
| **Leakage** | Career totals removed; per-game and ratios (debut/3-year windows) used so predictors don't encode career length. |

# Data Pipeline

-   **Script:** `00_nfl_EDA.Rmd` containing full cleaning, formulas, and visuals.
-   **Final Output:** `nfl/data/cleaned/5_ALL_DATA.csv`.
-   **Intermediates:** `1_cleaned_nfl_data.csv`, `2_reorder_nfl_data.csv`, `3_cleaned_nfl_debut.csv`, `4_cleaned_nfl_3year.csv`.
