---
title: "Data Documentation"
subtitle: "NFL Quarterback Retirement Research"
date: "2026-02-25"
author: "Emma Nguyen"
format: 
  pdf
html-math-method: mathjax
execute:
  eval: TRUE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
```

# Research Overview

**Purpose:** Survival Analysis of Quarterback retirement in National Football League (NFL).

**Built by:** `00_nfl_EDA.Rmd`

-   **Event:** Retirement = last season in NFL (no games after season *t*).
-   **Time:** Seasons since debut (`Career_Time`).
-   **Censoring:** Active through end of data → `Retired = 0`. Data **stops at 2016**; \~98 QBs censored (EDA).

**Raw inputs:** Basic_Stats.csv, Game_Logs_Quarterback.csv, Career_Stats_Passing.csv

-   Basic_Stats.csv:

-   Game_Logs_Quarterback.csv:

-   Career_Stats_Passing.csv:

**Sources:** [Kaggle NFL Statistics](https://www.kaggle.com/datasets/kendallgillies/nflstatistics/data) (scraped from official [NFL](http://www.nfl.com) website)

# Dataset:

**File:** `5_ALL_DATA.csv`

**Unit:** Each row is one QB (Quarterback). **Scope:** QBs with ≥1 row in Game_Logs_Quarterback. **Merge:** Career + Debut-Year + First-Three-Year Summary Statistics

Note: No raw career totals used as predictors to avoid leaking survival time.

## Load Data and Dimensions

```{r load data}
nfl_data <- read_csv(
  here("nfl", "data", "cleaned", "5_ALL_DATA.csv"),
  show_col_types = FALSE)

nfl_data
# 35 columns, 466 rows (Quarterbacks)
dim(nfl_data)
```

# Variable Reference

### *Identifiers and Outcome*

+-------------+-----------+---------------------------------------------------------------------------+
| Variable    | Type      | Description                                                               |
+=============+===========+===========================================================================+
| `Player_Id` | character | Unique Player ID.                                                         |
+-------------+-----------+---------------------------------------------------------------------------+
| `Name`      | character | Player name.                                                              |
+-------------+-----------+---------------------------------------------------------------------------+
| `Retired`   | 0/1       | **Outcome.** 1 = last season \< 2016; 0 = censored (active through 2016). |
+-------------+-----------+---------------------------------------------------------------------------+

### *Demographics and Career Frame*

+-------------------+------------+-------------------------------------------------------------------------------+
| Variable          | Type       | Description / formula                                                         |
+===================+============+===============================================================================+
| `Debut_Age`       | numeric    | Age at debut: `Age - (2016 - First_Year)` (Original Data: Age = age in 2016). |
+-------------------+------------+-------------------------------------------------------------------------------+
| `Height (inches)` | numeric    | Height of Player in inches.                                                   |
+-------------------+------------+-------------------------------------------------------------------------------+
| `Weight (lbs)`    | numeric    | Weight of Player in pounds.                                                   |
+-------------------+------------+-------------------------------------------------------------------------------+
| `First_Year`      | numeric    | Debut year (\~1970–2016).                                                     |
+-------------------+------------+-------------------------------------------------------------------------------+
| `Career_Time`     | numeric    | **Survival time:** seasons from debut = `Last_Year - First_Year + 1`.         |
|                   |            |                                                                               |
|                   |            | Ex. 11 seasons - 2001 to 2011                                                 |
+-------------------+------------+-------------------------------------------------------------------------------+

### *Game counts (ONLY Regular-Season)*

| Variable           | Description                     |
|--------------------|---------------------------------|
| `Career_Games`     | Total Games Played in Career.   |
| `Debut_Year_Games` | Total Games in debut season.    |
| `Three_Year_Games` | Total Games in first 3 seasons. |

### *Derived ratios (Career\_\*\| Debut\_\* \| Three_Year\_\*)*

-   **TD_INT:** `TD Passes / max(Ints, 1)` - Touchdown to Interception ratio; Infer accuracy/efficiency; higher → longer career (EDA).
-   **RUN_PASS:** `Rushing Yards / max(Passing Yards, 1)` — mobility; can be \<0 (sacks); higher → more hits, earlier retirement (EDA).

### *Completion and Per-Game stats (\*\_PG)*

-   **Completion_Percentage** (Career / Debut / Three-Year): Number of Throws Completions / Attempts; Infer accuracy/efficiency
-   **Per-Game Statistics:** `*_PG`, `Debut_*_PG`, `Three_Year_*_PG` for Yards (passing), Sacks, TD (touchdowns), Ints (interceptions), Rush (rushing yards).
    -   **Denominator** = Career_Games, Debut_Year_Games, or Three_Year_Games;
        -   **NA** when denominator = 0.

# Missingness and Ranges

**Missingness:** Debut stats have more NAs (when there are no debut games or 0 attempts). Three-year window has fewer NAs.

-   Most missingness is 'Debut_Completion_Percentage' with 181 of 466 rows NA.

```{r NAs}
missingness <- colSums(is.na(nfl_data))
missingness[missingness > 0]
max(missingness) 
```

**Ranges:**

-   First_Year (Debut Year): 1970–2016

-   RUN_PASS can be negative

-   Many zeros in TD_INT / per-game for short careers. Consider filtering e.g. `Big_T > 2` or `Total_Games > 0` for modeling.

# Data Structure and Summary

```{r summary}
glimpse(nfl_data)
summary(nfl_data)
```

# Limitations

+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| Topic          | Note                                                                                                                                  |
+================+=======================================================================================================================================+
| **Retirement** | Inferred (no games after season *t*). Misclassification possible (injury, etc.).                                                      |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| **Censoring**  | Cutoff 2016; QBs still active if Retired = 0. Consider sensitivity to cutoff.                                                         |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| **Leakage**    | Remove career totals (there for reference); per-game and ratios (debut/3-year windows) used so predictors don't encode career length. |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+

# Data Pipeline

-   **Script:** `00_nfl_EDA.Rmd` containing full cleaning, formulas, and visuals.
-   **Final Output:** `nfl/data/cleaned/5_ALL_DATA.csv`.
-   **Intermediates Files:** `1_cleaned_nfl_data.csv`, `2_reorder_nfl_data.csv`, `3_cleaned_nfl_debut.csv`, `4_cleaned_nfl_3year.csv`.
